OPP Veil of Darkness Report

### 2018-12-12 vodReport.Rmd
update filenames add abstract

data.cities <- read_rds(here::here("cache", "vod_full_data.Rds"))
data.states <- read_rds(here::here("cache", "vod_statewide_full_data.Rds"))
```

@@

# Summary of the dataset
We consider both city-level stops and statewide highway patrol stops. As in G & R, the analysis is limited to black and white drivers. Throughout the report, the term "minority driver" always refers to black drivers. The requirements of the test are quite stringent: the data must contain date, time and location of the stop, and the race of the driver. An additional control considered is geographic subdivision. The analysis is ultimately restricted to 7 cities and 10 states with all of these variables present.[^1]

[^1]: There are 26 cities with at least 75% non-`NA` values for date, time, driver race and stop location. 9 cities (Dallas, El Paso, Green Bay, Camden, Bakersfield, Little Rock, Plano, Hartford and Owensboro) are removed due to data issues, such as duplication, insufficient data in the intertwilight zone, or a negligible number of stops of minority drivers. Another 10 cities (Mesa, San Jose, Aurora, Wichita, Saint Paul, Grand Forks, Oklahoma City, Tulsa, San Antonio and Burlington) are removed because they do not provide one of region, precinct, reporting_area, district, beat, sector or police_grid_number. Though not included in this report due to time constraints, it may be interesting to run the regression that does not adjust for geographic subdivision on a larger dataset that includes these latter 10 cities.

@@

## Data Processing
As emphasised by Grogger & Ridgeway, racial driving patterns may vary significantly with time, so the analysis is limited to the “intertwilight zone”, which is the period of time during which it is light at least one day of the year and dark at least one day of the year. Since the darkness between sunset and dusk is ambiguous, “darkness” is defined as after dusk and “lightness” is defined as before sunset. Thus the intertwilight zone is defined as the minutes between the earliest dusk and the latest sunset. In addition, any stops between sunset and dusk are excluded due to the ambiguity.

@@

data.cities %>% 
  group_by(city) %>% 
  summarise(
    n_total = n(),
    min_year = min(year(date)), 
    max_year = max(year(date)),
    year_range = glue::glue("{min_year}-{max_year}"),
    percentage_minority = scales::percent(sum(is_minority_demographic) / n_total)
  ) %>% 
  select(-min_year, -max_year) %>% 
  arrange(desc(n_total)) %>% 
  knitr::kable(col.names = col_names, caption = "Summary of stops (city-level)")
```

@@

data.states %>% 
  group_by(state) %>% 
  summarise(
    n_total = n(),
    min_year = min(year(date)), 
    max_year = max(year(date)),
    year_range = glue::glue("{min_year}-{max_year}"),
    percentage_minority = scales::percent(sum(is_minority_demographic) / n_total)
  ) %>% 
  select(-min_year, -max_year) %>% 
  arrange(desc(n_total)) %>% 
  knitr::kable(caption = "Summary of stops (statewide)", col.names = col_names)
```

The calculation of sunset and dusk time requires latitude and longitude of the stop and the corresponding timezone. The raw data rarely includes this data, so latitude and longitude are estimated. For the city data, latitude and longitude are estimated using Google Maps, based on the officer’s description of the stop location. This procedure invariably makes a few errors, so any points more than 0.5 degrees in magnitude from the median latitude and longitude are considered spurious outliers and are dropped from the dataset. For the statewide data, the county centroid is used[^2].

[^2]: Shapefiles are used from the IPUMS NHGIS American Community Survey, 2016. Source: Steven Manson, Jonathan Schroeder, David Van Riper, and Steven Ruggles. IPUMS National Historical Geographic Information System: Version 13.0 [Database]. Minneapolis: University of Minnesota. 2018. http://doi.org/10.18128/D050.V13.0

We define geographic subdivisions to be use as a control in the regression analyses. For the statewide data, county is the appropriate subdivision, however the most appropriate subdivision varies by city. The subdivision selected for each city, one of precinct, district, beat or sector, is the geographic subdivision variable that has between 5 and 30 distinct values available in the data.

@@

A concern raised by G & R is the role of seasonal effects. To control for seasonal effects we fit a third model on the full dataset with an additional additive control that treats the month of the year as a categorical variable. In addition, we fit the time and geography-controlled regression on subsets of the data that includes only the two weeks before and after the two annual Daylight Savings Time (DST) shifts.[^3]

[^3]: The model fit with a month control, and the DST models are only fit on the city-level data, due to time constraints and the large size of the aggregated statewide dataset, however the code is readily available to fit these models on the statewide data.

@@

G & R define the veil of darkness parameter $K_{vod} = \frac{P(B|S, d = 0)P(\bar{B} | S, d = 1)}{P(\bar{B}|S, d = 0)P(B | S, d = 1)}$, where $B$ is the event a driver is black, $S$ is the event a driver is stopped, and $d$ is darkness as we define it, where $d = 1$ indicates it is dark and $d = 0$ indicates it is light. Theier analysis focuses on $log K$, which is negative in the presence of discrimination. As described in the paper, $-\beta_1$, where $\beta_1$ is the coefficient on the darkness variable in the logistic regression models, is an estimate for $logK$.  In this analysis we choose to consider the coefficient directly, as we find this a more intuitive and interpretable approach. A negative coefficient $\beta_1$ indicates that darkness is associated with a lower proportion of minorities stopped, and thus a negative coefficient is indicative of VOD discrimination.

@@

Figure 1 suggests that, between 6:30 and 7:30 PM, the proportion of minority drivers in the population of stopped drivers is lower after dark than before dark. The percentage of drivers stopped between 90 and 85 minutes before sunset is between 0.5 and 3 percentage points higher than the percentage of minority drivers amongst those stopped between 90 and 105 minutes after sunset.[^4]

[^4]: It should be noted that if we calculate the proportion of minorities by city (or geographic subdivision) and plot the mean of these proportions for each clock time and time relative to sunset, we produce a plot with very slightly positively sloped lines. This is likely due to the increased influence of smaller cities, which, as we see in Figure 2, tend to show less evidence of discrimination. We felt that Figure 1 is more representative of the effect present in the regressions included in Table 5, as it effectively weights the cities by their size, however, this phenomenon is worth further investigation.

@@

The lines in Figure 4 are sloping slightly downwards in the direction of the days where it is dark at the time of the stops. This suggests there may be VOD discrimination apparent in the data.[^5]

[^5]: See 4. The same argument applies in this case.

@@

by_city_dst %>%  
  mutate(
    low = estimate - 1.96*std.error, 
    high = estimate + 1.96*std.error,
    season = str_to_title(season)
  ) %>% 
  ggplot(aes(reorder(city, estimate), estimate)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_point() +
  geom_errorbar(aes(ymin = low, ymax = high)) +
  theme(
    axis.text.x = element_text(hjust = 1, angle = 30)
  ) +
  labs(
    x = NULL,
    y = "Coefficient Estimate with 95% CI",
    title = "Figure 5",
    subtitle = "The clock-time and geography adjusted models fit on the data in the two weeks\neither side of the DST shift do not show statistically significant effects at the 0.05 level for any city."
  ) +
  facet_grid(season ~ .)
```

@@

```{r statewide_models}
statewide_models <- 
  list(
    time_only = read_rds(here::here("cache", "vod_statewide_time_const.rds")),
    time_geo = read_rds(here::here("cache", "vod_statewide_geo_adj.rds"))
  )

@@

The lack of statistical significance in the DST analysis when adjusting for geographic subdivision is likely indicative that the restriction of the data to only four weeks of the year significantly reduces the power of the test. The coefficient estimates in the models fit only on the DST shift data look qualitatively similar to those in the models fit on all of the data, and the model fit on all the data with a control for month showed no significant difference from those that do not control for month. This provides further reason to believe that the discrepancy in statistical significance is a symptom of the significant reduction in the size of the datasets. The quantity of data could be increased by including cities that do not have subdivision data available and running only the clock-time adjusted regression.[^6] It will also be interesting to see whether the DST analysis on the much larger statewide dataset has the same issues of power.

[^6]: See 1 for the list of these cities.

@@

>>>>

data.cities <- read_rds(here::here("cache", "aggregated_cities_wsunset_and_subgeo.Rds"))
data.states <- read_rds(here::here("cache", "statewide_wsunset.Rds"))
```

# Abstract
In this report we implement the Veil of Darkness test (Grogger & Ridgeway, 2006) with a city-level dataset containing `r format(nrow(data.cities), big.mark = ",")` traffic stops across seven cities, and with a statewide dataset containing `r format(nrow(data.states), big.mark = ",")`  stops across ten states. The report is intended for readers who are relatively familiar with the test, so the explanation of its details is brief. Using logistic regression, we find statistically significant evidence of discrimination at the nationwide level in both datasets when controlling for the time and location of the stop. City-specific analysis identifies evidence of discrimination in Madison, WI; San Francisco, CA; Nashville, TN; and Philadelphia, PA. State-specific analysis __TODO__.

@@

We begin with a brief summary of the city-level and statewide datasets. This is followed by a description of the methods for data processing, implementation of the test, and analysis of results. Visualizations and regression results are presented first for the city-level data, followed by the statewide data. Finally, we address limitations of the test and potential future directions.

# Summary of the dataset
We consider both city-level stops and statewide highway patrol stops. As in G & R, the analysis is limited to black and white drivers. Throughout the report, the term "minority driver" always refers to black drivers. The requirements of the test are quite stringent: the data must contain date, time and location of the stop, and the race of the driver. An additional control considered is geographic subdivision. Requiring at least 75% non-`NA` values for date, time, driver race, location and geographic subdivision restricts the analysis to to 7 cities[^1]: Arlington, TX; Columbus, OH; Madison, WI; Nashville, TN; New Orleans, LA; Philadelphia, PA; and San Francisco, CA; and 10 states[^2]: Arizona; Connecticut; Michigan; Montana; North Dakota; Ohio; Tennessee; Texas; Wisonsin; and Wyoming.

[^1]: There are 26 cities with at least 75% non-`NA` values for date, time, driver race and stop location. 9 cities (Dallas, El Paso, Green Bay, Camden, Bakersfield, Little Rock, Plano, Hartford and Owensboro) are removed due to data issues, such as duplication, insufficient data in the intertwilight zone, or a negligible number of stops of minority drivers. Another 10 cities (Mesa, San Jose, Aurora, Wichita, Saint Paul, Grand Forks, Oklahoma City, Tulsa, San Antonio and Burlington) are removed because they do not provide one of region, precinct, reporting_area, district, beat, sector or police_grid_number. See Appendix 1 for analysis including the latter 10 cities.

[^2]: Data exploration has not been as rigorously performed on the statewide data. It is possible that some states or years should be removed due to data quality issues or very small percentages of minority drivers.

@@

## Data Processing
As emphasised by G & R, racial driving patterns may vary significantly with time, so the analysis is limited to the “intertwilight zone”, which is the period of time during which it is light at least one day of the year and dark at least one day of the year. Since the darkness between sunset and dusk is ambiguous, “darkness” is defined as after dusk and “lightness” is defined as before sunset. Thus the intertwilight zone is defined as the minutes between the earliest dusk and the latest sunset. In addition, any stops between sunset and dusk are excluded due to the ambiguity.

@@

data.cities %>% 
  group_by(city) %>% 
  summarise(
    n_total = n(),
    min_year = min(year(date)), 
    max_year = max(year(date)),
    year_range = glue::glue("{min_year}-{max_year}"),
    percentage_minority = scales::percent(sum(is_minority_demographic) / n_total)
  ) %>% 
  select(-min_year, -max_year) %>% 
  arrange(desc(n_total)) %>% 
  knitr::kable(col.names = col_names, caption = "Summary of stops in the intertwilight zone (city-level)")
```

@@

data.states %>% 
  group_by(state) %>% 
  summarise(
    n_total = n(),
    min_year = min(year(date)), 
    max_year = max(year(date)),
    year_range = glue::glue("{min_year}-{max_year}"),
    percentage_minority = scales::percent(sum(is_minority_demographic) / n_total)
  ) %>% 
  select(-min_year, -max_year) %>% 
  arrange(desc(n_total)) %>% 
  knitr::kable(caption = "Summary of stops in the intertwilight zone (statewide)", col.names = col_names)
```

The calculation of sunset and dusk time requires latitude and longitude of the stop and the corresponding timezone. The raw data rarely includes this data, so latitude and longitude are estimated. For the city data, latitude and longitude are estimated using Google Maps, based on the officer’s description of the stop location. This procedure invariably makes a few errors, so any points more than 0.5 degrees in magnitude from the median stop latitude and longitude are considered spurious outliers and are dropped from the dataset. For the statewide data, the county centroid is used[^3].

[^3]: Shapefiles are used from the IPUMS NHGIS American Community Survey, 2016. Source: Steven Manson, Jonathan Schroeder, David Van Riper, and Steven Ruggles. IPUMS National Historical Geographic Information System: Version 13.0 [Database]. Minneapolis: University of Minnesota. 2018. http://doi.org/10.18128/D050.V13.0

We define geographic subdivisions to be used as a control in the regression analyses. For the statewide data, county is the appropriate subdivision, however the most appropriate subdivision varies by city. The subdivision selected for each city, one of precinct, district, beat or sector, is the geographic subdivision variable that has between 5 and 30 distinct values available in the data.

@@

A concern raised by G & R is the role of seasonal effects. To control for seasonal effects we fit a third model on the full dataset with an additional additive control that treats the month of the year as a categorical variable. In addition, we fit the time and geography-controlled regression on subsets of the data that includes only the two weeks before and after the two annual Daylight Savings Time (DST) shifts.[^4]

[^4]: The model fit with a month control, and the DST models are only fit on the city-level data, due to time constraints and the large size of the aggregated statewide dataset, however the code is readily available to fit these models on the statewide data.

@@

G & R define the veil of darkness parameter $K_{vod} = \frac{P(B|S, d = 0)P(\bar{B} | S, d = 1)}{P(\bar{B}|S, d = 0)P(B | S, d = 1)}$, where $B$ is the event a driver is black, $S$ is the event a driver is stopped, and $d$ is darkness as we define it, where $d = 1$ indicates it is dark and $d = 0$ indicates it is light. Their analysis focuses on $log K$, which is negative in the presence of discrimination. As described in the paper, $-\beta_1$, where $\beta_1$ is the coefficient on the darkness variable in the logistic regression models, is an estimate for $logK$.  In this analysis we choose to consider the coefficient directly, as we find this a more intuitive and interpretable approach. A negative coefficient $\beta_1$ indicates that darkness is associated with a lower proportion of minorities stopped, and thus a negative coefficient is indicative of VOD discrimination.

@@

Figure 1 suggests that, between 6:30 and 7:30 PM, the proportion of minority drivers in the population of stopped drivers is lower after dark than before dark. The percentage of drivers stopped between 90 and 85 minutes before sunset is between 0.5 and 3 percentage points higher than the percentage of minority drivers amongst those stopped between 90 and 105 minutes after sunset.[^5]

[^5]: It should be noted that if we calculate the proportion of minorities by city (or geographic subdivision) and plot the mean of these proportions for each clock time and time relative to sunset, we produce a plot with very slightly positively sloped lines. This is likely due to the increased influence of smaller cities, which, as we see in Figure 2, tend to show less evidence of discrimination. We felt that Figure 1 is more representative of the effect present in the regressions included in Table 5, as it effectively weights the cities by their size, however, this phenomenon is worth further investigation.

@@

The lines in Figure 4 are sloping slightly downwards in the direction of the days where it is dark at the time of the stops. This suggests there may be VOD discrimination apparent in the data.[^6]

[^6]: See 5. The same argument applies in this case.

@@

by_city_dst %>%  
  mutate(
    low = estimate - 1.96*std.error, 
    high = estimate + 1.96*std.error,
    season = str_to_title(season)
  ) %>% 
  ggplot(aes(reorder(city, estimate), estimate)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_point() +
  geom_errorbar(aes(ymin = low, ymax = high)) +
  theme(
    axis.text.x = element_text(hjust = 1, angle = 30)
  ) +
  labs(
    x = NULL,
    y = "Coefficient Estimate with 95% CI",
    title = "Figure 5",
    subtitle = "The clock-time and geography adjusted models fit on the data in the two weeks\neither side of the DST shift do not show statistically significant effects\nat the 0.05 level for any city."
  ) +
  facet_grid(season ~ .)
```

@@

```{r statewide_models}
statewide_models <- 
  list(
    time_only = read_rds(here::here("cache", "vod_statewide_time_const.Rds")),
    time_geo = read_rds(here::here("cache", "vod_statewide_geo_adj.Rds"))
  )

@@

The lack of statistical significance in the DST analysis when adjusting for geographic subdivision is likely indicative that the restriction of the data to only four weeks of the year significantly reduces the power of the test. The coefficient estimates in the models fit only on the DST shift data look qualitatively similar to those in the models fit on all of the data, and the model fit on all the data with a control for month showed no significant difference from those that do not control for month. This provides further reason to believe that the discrepancy in statistical significance is a symptom of the significant reduction in the size of the datasets. The quantity of data could be increased by including cities that do not have subdivision data available and running only the clock-time adjusted regression, or adjusting for geography using the city rather than the subdivision.[^7] It will also be interesting to see whether the DST analysis on the much larger statewide dataset has the same issues of power.

[^7]: See 1 for the list of these cities. See Appendix 1 for preliminary results of this analysis.

@@

# Acknowledgements

My sincere thanks to Sharad for the opportunity to work on this project. It has been a steep but incredibly valuable learning curve that I am certain will serve me extremely well in the future. I have learned so much not only working with the largest dataset I've ever touched, but also being surrounded by such experienced and intelligent people who have been able to teach me how to ask the right questions and demand rigor. I would especially like to thank Amy and Dan for their generous support and patience throughout the quarter. Amy, in particular, has given so much of her time to answer my endless questions, help me through countless bugs and all without ever making me feel silly or like I'm wasting her time. She is an invaluable role model and I can't thank her enough.

# Appendix 1: Preliminary Analysis with 17 Cities

As mentioned in Footnote 1, there are 10 cities that were excluded from the above analysis because they do not include the geographic subdivision of the stops. As we point out above, these cities are theoretically eligible for analysis controlling for geography using the city, rather than the subdivision. In this appendix we provide __very preliminary__ results across the 17 cities. It shoud be recognized that the 10 cities were excluded from the above dataset _before_ exploration of data quality, and thus further exploration should be performed to ensure that these cities should all be included (for example, some cities have very few intertwilight stops).

Table 9 presents the regression estimates for the models that control for clock time, clock time and city, and clock time, city and month, fit on all stops from the 17 cities.

```{r extra_cities_reg}
read_rds(here::here("cache", "vod_no_geo_results.Rds")) %>% 
  mutate(controls = c("Clock time", "Clock time and City", "Clock time, City and Month")) %>% 
  knitr::kable(
    caption = "Logistic Regression Estimates (All stops; 17 Cities)", 
    col.names = c("Controls", "Coefficient Estimate", "Standard Error")
  )
```

Although the coefficient in the model that controls only for clock time is positive and statistically significant, the coefficients in the models that control for city are negative and statistically significant at the 0.05 level. The results when controlling for city match the intuition based on the regression with only 7 cities that controls for subdivision.

```{r, warning=FALSE}
dst_fall <- read_rds(here::here("cache", "vod_dst_no_geo_fall.Rds"))
dst_spring <- read_rds(here::here("cache", "vod_dst_no_geo_spring.Rds"))

tribble(
  ~season, ~adjustments, ~estimate,  ~se, 
  "Fall", "Clock time", coef(dst_fall$model_time_const)[2], coef(summary(dst_fall$model_time_const))[2, 2], 
  "", "Clock time and City", coef(dst_fall$model_geo_adjusted)[2], coef(summary(dst_fall$model_geo_adjusted))[2,2],
  "Spring", "Clock time", coef(dst_spring$model_time_const)[2], coef(summary(dst_spring$model_time_const))[2, 2],
  " ", "Clock time and City",  coef(dst_spring$model_geo_adjusted)[2], coef(summary(dst_spring$model_geo_adjusted))[2,2]
) %>% 
  column_to_rownames("season") %>% 
  knitr::kable(
    caption = "Logistic Regression Estimates (DST period; 17 cities)",
    col.names = c("Controls", "Coefficient", "Standard Error"), 
    row.names = TRUE
  )
```

When limiting the analysis to the DST period only, the clock time and city adjusted models no longer give statistically significant coefficient estimates. The models that adjust only for clock time still give positive, statistically significant coefficients.

# Appendix 2: Code Summary

The bulk of the veil of darkness code is implemented in `opp/lib/vod/veil_of_darkness_test.R`. The function `add_sunset_times` allows the user to filter to the intertwilight zone and save the results. Since adding sunset times is a long process, it's highly recommended to call `add_sunset_times` once, save the result, then use this result in multiple calls to `veil_of_darkness_test`. Set the boolean argument `has_sunset_times` to `TRUE` if you want to run `veil_of_darkness_test` on a pre-processed tibble.

The most important assumption made by `add_sunset_times` is that the data has a column `subgeography`, which is the grouping variable used to calculate timezones and sunset times. The function calculates __one__ timezone per `subgeography`, __one__ sunset time and __one__ dusk time per combination of `subgeography` and `date`, and __one__ intertwilight zone per `subgeography`.

The argument `has_geo_control` to `veil_of_darkness_test` indicates whether or not to run an additional model with an additive control for geographic subdivision. If `has_geo_control = TRUE`, then your data __must__ contain a column `geo_control`, which is used as this control. This allows you to change the variable used as the geographic control by creating a column called `geo_control` (using a `mutate`, `unite`, etc.).

The code to generate the data and results used in this report is contained in `opp/lib/vod/run_vod.R`. __WARNING__: This script takes many days to run if you want to generate all the data and run all the analyses. In particular, the statewide data is huge and the function `run_state_analysis()` will take days. Make sure you comment out the parts of the script you don't want to run.

As a caveat, I generated the data and results included in this report over time and over several smaller scripts, and created the master `run_vod.R` script in retrospect. I have done my best to debug by sight, but since the script takes days, I haven't actually tested `run_vod.R`. I don't guarantee against small bugs.
